Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-09-30 demystify.tex
updates

    \item \textbf{Not customizing the email}. These emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information. 2--3 paragraphs are enough.  Personally I also do not want to see course transcript or CV in the first email.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Employment:} You are allowed to work on-campus for up to \textbf{20 hours per week during the academic year} (because you still need to take classes) and \textbf{full-time} (typically 40 hrs) during official school breaks (e.g., summer and winter breaks). Off-campus employment requires authorization, which can be obtained through programs like Curricular Practical Training and Optional Practical Training described below.

@@

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation (which is typically way harder and takes much longer time than coursework).

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (\autoref{sec:funding}).

@@

\subsection{Funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded through either RA or TA.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as an RA. Moreover, some schools do have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has quite a few TAs for MS students; my courses in the past two years have MS TAs.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at Mason and UNL for many years. This allows me to have a good understanding of the process, the challenges students face, and what faculty are looking for. I also serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with MS admission---which is quite different than PhD). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

>>>>

    \item \textbf{Not customizing the email}. These emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information. 2--3 paragraphs are enough.  Personally I also do not want to see course transcript or CV in the first email.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
    \item \textbf{Not using plain text}. Fancy format, colors, or fonts are not professional. CS people like plain text.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Employment:} You are allowed to work on-campus for up to \textbf{20 hours per week during the academic year} (because you still need to take classes) and \textbf{full-time} (typically 40 hrs) during official school breaks (e.g., summer and winter breaks). Off-campus employment requires authorization, which can be obtained through CPT and OPT programs described below.

@@

  \item Course requirements: MS has a specific number of courses (typically can be done in 2 years). You graduate with an MS when you're done with the courses. PhD also has coursework requirements (typically taken in your first 2 years) but focuses on research after that.  You graduate with a PhD when you have done enough research and written a dissertation, which usually takes much longer time than coursework.

@@

  \item Funding: MS is typically \emph{not funded}, while PhD is typically funded (\autoref{sec:funding}). See \autoref{sec:ms-funding} for more details on MS funding.

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded through either RA or TA.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

However, there are always exceptions, e.g., if you have a strong background and can demonstrate that you can be productive, then a prof. might be willing to take you as their RA.
Moreover, some schools have TAs for MS students, and you can apply for these positions. For example, Mason CS dept has quite a few TAs for MS students; my courses in the past two years have MS TAs.

@@

\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). I have been in the PhD admission process at Mason and UNL for many years. This allows me to have a good understanding of the process, the challenges students face, and what faculty are looking for. I also serve as the program director of the MS program in Software Engineering at Mason (thus also have experience with MS admission---which is quite different than PhD as discussed in \autoref{chap:ms}). My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.
